Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

The Wilkins-Diehr paper

>>>>

The Wilkins-Diehr paper describes an NSF software institute effort to build a community of those creating science gateways or
web portals for science. These gateways, as described in some of the other papers in this section, can be quite capable, having a 
remarkable impact on science.

This paper mentioned challenges highlighted by other papers in this section - mainly the conflict between funding for research
vs infrastructure and the challenges around getting academic credit for infrastructure. Because the authors have studied
many projects, they've also observed how development is often done in an isolated hobbyist environment. Developers are unable
to take advantage of similar work done by others. isolation even when projects have common needs.
But often projects struggle for sustainable funding because they provide infrastructure to conduct
research and many times only the research is funded. Gateways also may start as small group research project, taking off in popularity
once they go live without any long term plans for sustainability or without resources in the project to plan for such.
Subsequent disruptions in service can limit effectiveness and test the limits of the research community's trust. The impact
of gateways can be increased substantially if we understand what makes them successful.

Recommendations from an early study of successful gateways include the following. Leadership and management teams should 
design governance to represent multiple strengths and perspectives, plan for change and turnover in the future, 
recruit a development team that understands both the technical and domain-related issues, consider sustainability and measure success early and often.
Successful projects recognize the benefits and costs of hiring a team of professionals, demonstrate credibility through stability and 
clarity of purpose, leverage the work of others and plan for flexibility.
Successful projects identify an existing community and understand that communities' needs before beginning. Projects adapt
as the communities' needs evolve. For funders, the lifecycle of technology projects must be considered. Solicitations should
be designed to reward effective planning, recognize the benefits and limitations of both technology innovation and reuse,
expect adjustments during the production process, copy effective models from other industries and sectors, and encourage partnerships that support gateway sustainability.

Through a business incubator type approach, the institute plans to provide a variety of services that could be shared amongst
projects. Consultation and resources on topics such as business plan development, software engineering practices, 
software licensing options, usability, security and project management as well as a software repository and hosting 
service will be available. Experts will be available for multi-month assignments to help research teams build their own gateways,
teaching them how to maintain and add to the work after the collaboration ends. Forums, symposia and an annual conference
will connect members of the development community.A modular, layered framework that supports community contributions and 
allows developers to choose components will be delivered and finally 
workforce development activities will help train the next generation for careers in this cross- disciplinary area and 
build pools of institutional expertise that many projects can leverage. Shared services and forums can dramatically reduce the cost
of building and sustaining gateways. Workforce development can encourage technology professionals to remain in the sciences.
